```{r set-options, echo=FALSE,comment=NA, cache=FALSE}
options(width=87)
```

2. Reworking and Recoding Data
=================

Often when working on a project you will have a data set that will contain additional information that you don't need for your analysis; or, have attributes which aren't specified as you require. This helpsheet explains how to remove and add additional attributes. 

For example, let's say we have a data set as follows:

Name | Age | Place | School | Degree
--- | --- | --- | --- | ---
John | 20 | Liverpool | Hillside High School | Geography BA (Hons)
Rachel | 21 | Norwich | Colman High School | Geography & Archaeology BA (Joint Hons)
... | ... | ... | ... | ...

And we are only interested in people's age for this exercise. As such, we don't need all of the other data. 

Before we start, we need to setup the working directory and read in the data:

>>*For more information on working directories, see the worksheet '1. R Basics'. Remember to create the folder `R work` if it doesn't exist already.*

```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/2-recoding-data")
```
<!---```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/worksheets/2-recoding-data")
``` -->
```{r,echo=FALSE,comment=NA,results='hide'}
  file_location <- "example.csv"
  data <- read.csv(file_location, header = TRUE)
```
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Read data from the web  
  data <- read.csv("http://data.alex-singleton.com/r-helpsheets/2/example.csv", header = TRUE)
```

We will now display this to check it has been read in correctly:

```{r,results='hide'}
  data
```

Which should give you this:

```{r,echo=FALSE,comment=NA}
  data
```

The `subset` command can be used to extract just the specified columns (and/or rows) from the data set. For example:

```{r,results='hide'}
  subset(data, select = c("Name", "Age"))
```
 
```{r,results='hide'}
  subset(data, Place == "Liverpool", select = c("Name", "Age"))
```

We can also store this as a new object:

```{r,results='hide'}
  data.Liverpool <- subset(data, Place == "Liverpool", select = c("Name", "Age"))
```

Because the statement assigns the output of the subset function to the new object called `"data.Liverpool"`, nothing will be printed. As such, we can check by typing `data.Liverpool`:

```{r,echo=FALSE,comment=NA}
  data.Liverpool
```

Adding a column to a data frame is done using the $ symbol. We will initially store `NA` (i.e. no value) in the column. 

```{r,results='hide'}
  data.Liverpool$diff100 <- NA
```

We also use the same principle to calculate the age difference from 100

```{r,results='hide'}
  data.Liverpool$diff100 <- 100 - data.Liverpool$Age
```

Perhaps we decide that we don't like the label of the first column "`Name`" and that it would be more appropriate to call it "`FirstName`". To make this change we create a variable with the column labels that we want:

```{r,results='hide'}
  new_column_names <- c("FirstName","Age","diff100")
```

When doing this it is always a good idea to check that the length of the object we have just created (it should be `3`) is the same as the number of columns in our data frame. 

```{r,results='hide'}
length(new_column_names)
```

```{r,results='hide'}
ncol(data.Liverpool)
```

We can then add the new column names to the data frame:

```{r,results='hide'}
colnames(data.Liverpool) <- new_column_names
```

Check the data frame now, and the names should be changed. 

```{r,echo=FALSE,comment=NA}
  data.Liverpool
```

Instead of recording people's age in years, perhaps we just need this in two categories - 21 and over, and under 21. We can _recode_ the `Age` variable into a new variable as follows:

```{r,results='hide'}
  data.Liverpool$AgeCat[data.Liverpool$Age < 21] <- "Under 21"
  data.Liverpool$AgeCat[data.Liverpool$Age >= 21] <- "21 or over"
```

This will create the new variable, `AgeCat`. To see what has happened to the object, print the `data.Liverpool` again:

```{r,echo=FALSE,comment=NA}
  data.Liverpool
```
