4. Joining Data
=====

When doing spatial data analysis, it is quite common to need to merge different data sets together. There are two main ways of doing this, firstly using the `merge()` command which will match attribute data in data frames, and secondly, using the match technique, which works with attribute data in shape files. 

### Merging Attribute Data in Data Frames

The `merge()` function allows us to take two data sets and combine them into one, based on a common variable. To test this, import the following data by running this command:

```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/4-attribute-joins")
  file_location <- "example.csv"
```
<!---
```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/helpsheets/4-attribute-joins")
```
-->
```{r,echo=FALSE,comment=NA,results='hide'}
  file_location <- "example.csv"
  data <- read.csv(file_location, header = TRUE, skip = 1)
```
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Read data from the web  
  data <- read.csv("http://data.alex-singleton.com/r-helpsheets/4/example.csv", header = TRUE, skip = 1)
```

```{r,results='hide',echo=FALSE,comment=NA}
data <- read.csv(file_location, header = TRUE, skip = 1)
```

And to check that it has imported correctly, which is always a good idea, run:

```{r,results='hide'}
  data
```

Which should output:

```{r,echo=FALSE,comment=NA}
  data
```

You now need to create another data frame which we will use as an example. You could create another csv file and import this; however, we will illustrate another way of achieving this by joining a series of vector lists.

```{r,results='hide'}
  # Create a person vector
  Person <- c("Paul", "Mike", "John", "Helen", "Mia", "Leo", "Rachel")
  # Create a favourite functions vector
  Function <- c("merge()", "read.csv()", "colnames()", "ncol()", "length()", "getwd()", "save.image()")
  # We can now join these two vectors into a new data frame of favourite functions
  fav_fun <- data.frame(Person, Function)
  # View the fav_fun
  fav_fun
```
Which should look like this:

```{r,echo=FALSE,comment=NA}
  fav_fun
```

We now have two data sets; `data`, which contains a list of people, locations and schools and `fav_fun`, which contains a list including those people as well as additional people who have attended R workshops. 

The next step is to combine the two. What we are going to do is select the people in the `fav_fun` data frame who also appear in the `test` data frame, and copy their favourite R function into a new data frame, along with all the information from `test`. 

We will refer to the two data frames as `x` and `y`. The x data frame is `data`; and the y is `fav_fun`. In `x`, the column containing the list of people is called "Name", and in `y`, it is called "Person". The parameters of the merge function first accept the two table names, and then the lookup columns as `by.x` or `by.y`. You should also include `all.x=TRUE` as a final parameter. This tells the function to keep all the records in `x`, but only those in `y` that match. 

```{r,results='hide'}
  People_And_Functions <- merge(data, fav_fun, by.x = "Name", by.y = "Person", all.x = TRUE)
```

To see what this command has done, type `People_And_Functions` to show the content of the new data frame. This should look like:

```{r,echo=FALSE,comment=NA}
  People_And_Functions
```

If the by column names were named the same in both `x` and `y` (e.g. both called "Name"), we could specify this more simply with `by="column name"` rather than `by.x` and `by.y`; and finally, a critical issue when making any join is assuring that the "`by`" columns are in the same format.

### Match data in a Shapefile

The `match()` function works in a very similar way to `merge()` but can be used to append attribute data to a  shape file. 'merge()' will often cause errors when working with spatial data frames.

Load the required packages and example shapefile from helpsheet "3. Importing External Data". 

```{r,results='hide',message=FALSE}
  library(rgdal)
```
```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/4-attribute-joins")
  file_location <- "example.csv"
```
<!---
```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/worksheets/4-attribute-joins")
```
-->
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Download data.zip from the web
  download.file("http://data.alex-singleton.com/r-helpsheets/4/data.zip", "data.zip")
  # Unzip file
  unzip("data.zip")
```
```{r,results='hide'}
  # Read in shape file
  Wards <- readOGR(".", "england_caswa_2001")
```

```{r,eval=FALSE,results='hide'} 
  # Plot Wards to check it has been imported correctly
  plot(Wards)
```

```{r,results='hide',echo=FALSE,comment=NA,warning=FALSE}
  pdf('plot1.pdf', 4, 4)
  plot(Wards)
  dev.off()
```

![Image](plot1.pdf)\
  
  
We now have the content of the `Wards` shapefile in R. Have a look at the content of the data in the data slot:

```{r,results='hide'}
  head(Wards@data)
```

```{r,echo=FALSE,comment=NA}
  head(Wards@data)
```

We are now going to append the following data onto it, which are index scores for the rate of diabetes prevelance:

Ward | Rate
--- | ---
00BYGC | 50
00BYFN | 198
00BYFU | 56
00BYFC | 78
00BYFG | 123
00BYFS | 21

Run this code to create this data frame:

```{r,results='hide'}
  # Create an ons code vector
  Ward <- c("00BYGC", "00BYFN", "00BYFU", "00BYFC", "00BYFG", "00BYFS")
  # Create a rate vector
  Rate <- c(50, 198, 56, 78, 123, 21)
  # We can now join these two vectors into a new data frame of wards_diabetes
  wards_diabetes <- data.frame(Ward, Rate)
  # View the wards_diabetes
  wards_diabetes
```
This should look like:

```{r,echo=FALSE,comment=NA}
  wards_diabetes
```

We can then use the `match()` function to append these diabetes rates on to `Wards@data`, by matching the `Ward` column from the `wards_diabetes` data frame to the `ons_label` column in the data slot of the wards SpatialPolygonsDataFrame.

```{r,results='hide'}
  Wards@data <- data.frame(Wards@data, wards_diabetes[match(Wards@data[, "ons_label"], wards_diabetes[, "Ward"]), ])
```

And to check, run:

```{r,results='hide'}
  head(Wards@data)
```

```{r,echo=FALSE,comment=NA}
  head(Wards@data)
```

We have now appended the data, but also have the ward listed twice. To remove this, run:

```{r,results='hide'}
  Wards@data$Ward <- NULL
  head(Wards@data)
```

Which changes `Wards@data` to:

```{r,echo=FALSE,comment=NA}
  head(Wards@data)
```
