# 5. Basic Maps

This helpsheet shows you how to make a simple map using the `GISTools` package. 

To start with, we need to load the `GISTools` package as well as some other packages we need:

```{r,results='hide', message=FALSE}
  library(rgdal)
  library(GISTools)
  library(RColorBrewer)
```

We also need to load a data set, which in this example, relate to Lower Layer Super Output (LSOA) zones within Liverpool, and also an outline of England. The following commands will set your working directory, download, unzip and load the data files. 

```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/5-basic-maps")
```
<!---
```{r,echo=FALSE,comment=NA,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/helpsheets/5-basic-maps")
```
-->
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Download data.zip from the web
  download.file("http://data.alex-singleton.com/r-helpsheets/5/data.zip", "data.zip")
  # Unzip file
  unzip("data.zip")
```
```{r,results='hide'}
  # Read in both shapefiles
  LSOA <- readOGR(".", "england_LSOA_2011_dwelling_count")
  outline <- readOGR(".", "England_ol_2011_gen_clipped")
```

We can do a very basic plot of the map using:

```{r,eval=FALSE,highlight=TRUE}
  plot(LSOA)
```

Which gives us a map, just showing the boundaries of the LSOAs. 
  
  
  
```{r,results='hide',echo=FALSE,comment=NA,warning=FALSE,fig.cap = "Map of Wards in Liverpool"}
  pdf('plot1.pdf', 5, 5)
  plot(LSOA)
  dev.off()
```

![Image](plot1.pdf)\
  
We can also plot an outline of England in a similar way.

```{r,eval=FALSE,highlight=TRUE}
plot(outline)
```

This replaces the first map, but we can get R to overlay one on top of the other, by using the command `add = TRUE`. The order of plots is key here - R will maintain the scale and extent of the first map. We can also adjust the colour of the border to a red colour (`border="red"`), and the fill colour (`col="#2C7FB820"`) a shade of blue. These represent two ways of specifying colours. The second contains eight alphanumerics, the first six relate to a HEX colour code. To view various colours that can be used in R, have a look at the website http://research.stowers-institute.org/efg/R/Color/Chart/ColorChart.pdf. The final two characters are the level of transparency (in this case 20%). _Sometimes when running R in Windows, the transparency option will not work - it will just fill it with a solid colour. In this case, just remove the `col = "#2C7FB820"` section from the plot command to just generate a red outline._


```{r,eval=FALSE,highlight=TRUE}
  # Plot the LSOA Map
  plot(LSOA)
  # Overplot the outline map
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
```

```{r,results='hide',echo=FALSE,comment=NA}
  pdf('plot2.pdf', 5, 5)
  plot(LSOA)
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
  dev.off()
```

\begin{center}
\includegraphics{plot2.pdf}
\par
\end{center}
    
The LSOA data frame contains some more information, which we can see by looking in the data slot of the object: 

```{r,results='hide'}
  head(LSOA@data)
```

```{r,echo=FALSE,comment=NA}
  head(LSOA@data)
```

This shows us that the shape file contains a field called '`COUNT_DWELL`' which contains the count of the number of dwellings in each LSOA. We can use this to create a choropleth map with:

```{r,eval=FALSE,highlight=TRUE}
  choropleth(LSOA, LSOA$COUNT_DWEL)
```

```{r,results='hide',echo=FALSE,comment=NA}
  pdf('plot3.pdf', 5, 5)
  choropleth(LSOA, LSOA$COUNT_DWEL)
  dev.off()
```

\begin{center}
\includegraphics{plot3.pdf}
\par
\end{center}
    
This map is ok, but we can easily make it more effective with a few extra commands. The new commands include:

1. `brewer.pal` which returns a set of colours from a range of pre-set palettes that look good on maps. In this case, we are getting `5` colours from the `"Blues"` palette. For more information on the R command, type '`?brewer.pal`' into R, for more information on the concept, see http://colorbrewer.org. 

1. `auto.shading` which categorises the data we want to show on to the map (in this case, `LSOA$COUNT_DWELL`) into the specified number of categories (`5`), coloured with the specified colours (`cols = brewer.pal(5, "Blues")`).

1. `choro.legend` and `north.arrow` both have a set of coordinates as one of their parameters (e.g. `331089, 384493`). These say where the object is located on the map. You may have to fiddle with these to get the spacing correct (see note below).

Run the commands below in R, and read the text below for more information. 

```{r,eval=FALSE,highlight=TRUE}
  # Set colour and number of classes
  shades <- auto.shading(LSOA$COUNT_DWEL, n = 5, cols = brewer.pal(5, "Blues"))
  # Draw the map
  choropleth(LSOA, LSOA$COUNT_DWEL, shades)
  # Add a legend
  choro.legend(331089, 384493, shades, fmt = "%g", title = "Count of Dwellings")
  # Add a title to the map
  title("Count of Dwellings by LSOA, 2011")
  # add Notth arrow
  north.arrow(332308, 387467, 300)
  # Draw a box around the map
  box(which = "outer")
```

```{r,results='hide',echo=FALSE,comment=NA}
  pdf('plot4.pdf', 7, 7)
  # Set colour and number of classes
  shades <- auto.shading(LSOA$COUNT_DWEL, n = 5, cols = brewer.pal(5, "Blues"))
  # Draw the map
  choropleth(LSOA, LSOA$COUNT_DWEL, shades)
  # Add a legend
  choro.legend(328089, 384493, shades, fmt = "%g", title = "Count of Dwellings")
  # Add a title to the map
  title("Count of Dwellings by LSOA, 2011")
  # add Notth arrow
  north.arrow(332308, 387467, 300)
  # Draw a box around the map
  box(which = "outer")
  dev.off()
```

See the next page for the map. 

You might find you will need to adjust the location or size of the legend to get this to fit onto your plot correctly. To find a new set of location coordinates, type `locator()` into the terminal and press enter. After doing this, when you hover over the plot, the mouse will turn into a cross. If you click, and then right-click and choose 'Stop', the location of the click is printed to the terminal - you can use these to re-position items in the plot.

To change the size of the legend, use the `cex = ` command. Update the `choro.legend` line to read `choro.legend(328089, 384493, shades, fmt = "%g", title = "Count of Dwellings", cex = 1.1)` and see what happens. The `cex` value is a multiple which increases or decreases the size of the legend. Experiment with this until you find something that works well. 

For more information on the `GISTools` package, have a look at http://cran.r-project.org/web/packages/GISTools/  GISTools.pdf. 

\begin{center}
\includegraphics{plot4.pdf}
\par
\end{center}
    
