# 6. Basic Spatial Analysis

This helpsheet will explore a variety of basic spatial analysis techniques, including *clipping*, *point in polygon* and *buffering*.

### Clipping

Clipping allows us to use one set of boundaries to cut another, a bit like using a cookie cutter.

```{r,results='hide',message=FALSE}
  # Load the Libaries
  library(rgdal)
  library(maptools)
  library(rgeos)
  library(stringr)
```

```{r,echo=FALSE,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/6-basic-spatial-analysis")
```
<!---
```{r,echo=FALSE,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/helpsheets/6-basic-spatial-analysis")
```
-->
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Download data.zip from the web
  download.file("http://data.alex-singleton.com/r-helpsheets/6/data.zip", "data.zip")
  # Unzip file
  unzip("data.zip")
```
```{r,results='hide'}
  # Read in both shape files
  LSOA <- readOGR(".", "england_LSOA_2011")
  outline <- readOGR(".", "England-outline")
```

First of all, we can plot the LSOA zones in Liverpool. 

```{r,results='hide',eval=FALSE}
  # Plot the LSOA Map
  plot(LSOA)
```

```{r,results='hide',echo=FALSE,warning=FALSE}
  pdf('plot1.pdf', 5, 5)
  plot(LSOA)
  dev.off()
```

![Image](plot1.pdf)\
  
    
    
We can also plot the England outline, but if we just run `plot(outline)` it will replace the LSOA plot on the display. To add the `outline` layer to the existing plot window, we can run the code below which will plot the outline with a red border and we can also adjust the colour of the border to a red colour (`border="red"`), and the fill colour (`col="#2C7FB820"`) a shade of blue. These represent two ways of specifying colours. The second contains eight alphanumerics, the first six relate to a HEX colour code. To view various colours that can be used in R, have a look at the website http://research.stowers-institute.org/efg/R/Color/Chart/ColorChart.pdf. The final two characters are the level of transparency (in this case 20%). _Sometimes when running R in Windows, the transparency option will not work - it will just fill it with a solid colour. In this case, just remove the `col = "#2C7FB820"` section from the plot command to generate a red outline._

```{r,results='hide',eval=FALSE}
  # Overplot the outline map
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
```

```{r,results='hide',echo=FALSE,warning=FALSE}
  pdf('plot2.pdf', 5, 5)
  plot(LSOA)
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
  dev.off()
```

\begin{center}
\includegraphics{plot2.pdf}
\par
\end{center}

As you will notice, the LSOA boundaries cross the River Mersey and stop at the river centre line. This doesn't look very nice, so we can tidy this up by getting R to clip the LSOA boundaries where they cross the England outline border. 

We do this using the `gIntersection` command, passing it the two layer variables (`outline` and `LSOA`). We can also tell R we just want it to use the area covering the LSOAs by specifying `byid = TRUE` and `id = my_area_id`. Be aware that the `gIntersection` command may take up to 90 seconds to run - do not worry if your computer appears to freeze. Just wait for the command to complete. 

```{r,results='hide',eval=FALSE}
  # set the area we want to cut
  my_area_id <- as.character(LSOA@data$ZONECODE)
  # run the Intersection command, saving output to clipLSOA, this may take anywhere up to 90 seconds to run
  clipLSOA <- gIntersection(LSOA, outline, byid = TRUE, id = my_area_id)
  #replot the map as above to see what we have done
  plot(clipLSOA)
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
```

```{r,results='hide',echo=FALSE,warning=FALSE}
  pdf('plot3.pdf', 5, 5)
  # set the area we want to cut
  my_area_id <- as.character(LSOA@data$ZONECODE)
  # run the Intersection command, saving output to clipLSOA, this may take a few seconds to run
  clipLSOA <- gIntersection(LSOA, outline, byid = TRUE, id = my_area_id)
  #replot the map as above to see what we have done
  plot(clipLSOA)
  plot(outline, add = TRUE, border = "red", col = "#2C7FB820")
  dev.off()
```

\begin{center}
\includegraphics{plot3.pdf}
\par
\end{center}

We have now removed the parts of the LSOAs that overlap the coastline, and the map looks much more attractive.

### Point in Polygon Analysis

Point in polygon analysis is useful when you want to create a subset of points from a larger set based on their spatial location. In this example we will load a list of locations that relate to all doctors surgeries in England, and use the polygons of ward boundaries in Leeds to create a subset of the Leeds doctors surgeries. To begin with, we need to load the libraries and get the GP and Wards data.

```{r,results='hide'}
  # Load the Libaries
  library(maptools)
  library(rgeos)
```

```{r,echo=FALSE,results='hide'}
  setwd("/Users/nickbearman/Dropbox/r-helpsheets/helpsheets/6-basic-spatial-analysis")
```
<!---
```{r,echo=FALSE,results='hide'}
  setwd("C:/Users/Nick & Louise/Dropbox/r-helpsheets/worksheets/6-basic-spatial-analysis")
```
-->
```{r,eval=FALSE,results='hide'}
  # Set working directory
  setwd("M:/R work")
  # Download data.zip from the web
  download.file("http://data.alex-singleton.com/r-helpsheets/6/data.zip", "data.zip")
  # Unzip file
  unzip("data.zip")
```
```{r,results='hide'}
  # Read in shapefile
  Wards <- readShapeSpatial("CAS-leeds", proj4string = CRS("+init=epsg:27700"))
```

It's worth having a quick look at the Leeds data so we know what it looks like:

```{r,eval=FALSE,results='hide'} 
  # Plot Wards to check it has been read in correctly
  plot(Wards)
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot4.pdf', 5, 5)
  plot(Wards)
  dev.off()
```

\begin{center}
\includegraphics{plot4.pdf}
\par
\end{center}

The doctors surgeries data is quite untidy - once we've read it in, we need to remove some extra columns that we don't need, and rename the ones we do.

```{r,results='hide',comment=NA}
  # Get Data
  GP <- read.csv("General Practices 2006.csv", header = TRUE, skip = 3)

  # Extract the columns we want
  GP <- subset(GP, select =c("Practice.Doctor.s.Name", "Easting", "Northing"))

  # Rename the columns to something more helpful
  colnames(GP) <- c("Surgery", "Easting", "Northing")
```

```{r,eval=FALSE,results='hide'} 
  # Do a plot to check what the data look like
  plot(GP$Easting, GP$Northing)
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot5.pdf', 6, 6)
  # Do a plot to check what the data look like
  plot(GP$Easting, GP$Northing)
  dev.off()
```

\begin{center}
\includegraphics{plot5.pdf}
\par
\end{center}

This should look like the above. The next stage is to convert the data into a SpatialPointsDataFrame. 

```{r,results='hide'}
  # Remove those GP without Easting or Northing
  GP <- subset(GP, Easting != "" & Northing != "")
  # Create a unique ID for each GP
  GP$GP_ID <- 1:nrow(GP)
  # Create the SpatialPointsDataFrame
  GP_SP <- SpatialPointsDataFrame(coords = c(GP[2], GP[3]), data = data.frame(GP$Surgery, GP$GP_ID), proj4string = CRS("+init=epsg:27700"))                                     
```

The first line contains a `subset` command which removes any of the entries which have a blank value for Northings or Eastings. `!=` means 'not equal to' and `&` means 'AND' so in "English" the command reads "overwrite the GP data frame with a subset of the GP data frame where the Easting field is not blank and the Northing field is not blank".

In a SpatialPointsDataFrame each entry must have a unique ID, so the second line creates an ID in the column `GP_ID`. The third line brings together the different elements to create the SpatialPointsDataFrame, `GP-SP`. `GP[2]` and `GP[3]` are the `Easting` and `Northing` columns respectively, and the `data =` section tells R which bits of the data frame to include. In this case we only want the surgery name (`GP$Surgery`) and the ID number (`GP$GP_ID`). The final term (`proj4string`) specifies which projection the data set is in - in this case, British National Grid (`epsg:27700`).

```{r,eval=FALSE,results='hide'} 
  # Show the results
  plot(GP_SP)       
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot6.pdf', 6, 6)
  # Show the results
  plot(GP_SP)       
  dev.off()
```

\begin{center}
\includegraphics{plot6.pdf}
\par
\end{center}

This plot will look similar to the previous one, but the data are now stored in a SpatialPointsDataFrame. We now can calculate a point in polygon, i.e. to select those points which lie within the boundary of Leeds. The forth line below uses a `!is.na` command. `is.an` is a command to test whether a value is 'NA' and `!` means the inverse, so the command is testing whether the value (of `GP_SP@data$label`) is not `NA`. 

```{r,results='hide'}
  # point in polygon - returns a dataframe of the attributes of the polygons
  # that the point is within.
  o <- over(GP_SP, Wards)

  # Many of these will be NA values - because most GPs are not in Leeds!
  head(o)                         

  # Add the attributes back onto the GP_SP SpatialPointsDataFrame (they are the same length)
  GP_SP@data <- cbind(GP_SP@data, o)

  # Use the NA values to remove those points not within Leeds
  GP_SP_Leeds <- GP_SP[!is.na(GP_SP@data$label), ]                       
```

```{r,eval=FALSE,results='hide'} 
  # Map your results
  plot(GP_SP_Leeds)        
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot7.pdf', 6, 6)
  # Map your results
  plot(GP_SP_Leeds)          
  dev.off()
```

\begin{center}
\includegraphics{plot7.pdf}
\par
\end{center}

We can also plot the points over the Leeds LSOAs:

```{r,eval=FALSE,results='hide'} 
  plot(Wards)
  plot(GP_SP_Leeds, add = TRUE)     
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot8.pdf', 6, 6)
  plot(Wards)
  plot(GP_SP_Leeds, add = TRUE)        
  dev.off()
```

\begin{center}
\includegraphics{plot8.pdf}
\par
\end{center}

We can also view the data in the `GP_SP_Leeds` data frame. 

```{r,results='hide'}
  # View the data slot of the results
  head(GP_SP_Leeds@data)
```

```{r,echo=FALSE,comment=NA}
  head(GP_SP_Leeds@data)
```


## Buffers

>> _This section looks at buffers. It carries on from the section on points in polygon, so make sure you complete that section first._

Buffers are often used in spatial analysis for defining context of points. In this example we will calculate a buffer from the doctors surgeries of a 10 minute walking distance, based on an average of 3 mph, which is around 1608m. 

The rgeos package has a function called `gBuffer()` that can be used to create buffers around points, lines or polygon objects. In the following example we create a new SpatialPolygons object called `GP_SP_Leeds_Buffers`. This then needs to be converted into a SpatialPolygonsDataFrame object by joining the `@data` from `GP_SP_Leeds` back onto `GP_SP_Leeds_Buffers`. Spatial Polygons objects do not have the data slot.

```{r,results='hide'}
  # buffers
  GP_SP_Leeds_Buffers <- gBuffer(GP_SP_Leeds, width = 1608, byid = TRUE)
  
  # Convert GP_SP_Leeds_Buffers into a SpatialPolygonsDataFrame (rather than
  # SpatialPolygons) by joining the data of the GP_SP_Leeds
  # SpatialPolygonsDataFrame
  GP_SP_Leeds_Buffers <- SpatialPolygonsDataFrame(GP_SP_Leeds_Buffers, GP_SP_Leeds@data)
```

We can also now plot this on top of the Wards map. 

```{r,eval=FALSE,results='hide'} 
  # Wards wards
  plot(Wards, axes = FALSE, col = "#6E7B8B", border = "#CAE1FF")
  # GP locations
  plot(GP_SP_Leeds, pch = 19, cex = 0.4, col = "#5CACEE", add = TRUE)
  # catchment buffers
  plot(GP_SP_Leeds_Buffers, axes = FALSE, col = NA, border = "red", add = TRUE) 
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  pdf('plot9.pdf', 6, 6)
  # Wards wards
  plot(Wards, axes = FALSE, col = "#6E7B8B", border = "#CAE1FF")
  # GP locations
  plot(GP_SP_Leeds, pch = 19, cex = 0.4, col = "#5CACEE", add = TRUE)
  # catchment buffers
  plot(GP_SP_Leeds_Buffers, axes = FALSE, col = NA, border = "red", add = TRUE)      
  dev.off()
```

\begin{center}
\includegraphics{plot9.pdf}
\par
\end{center}
